---
title: "LiDAR_01"
author: "Jing-Zhong Lu"
date: "2025-06-10"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#library(rgdal)
library(lidR)
library(viridis)
library(ggplot2)
library(terra)
library(patchwork)
library(gstat)
library(rgl)
library(sf)
```

## Materials
LidR: https://r-lidar.github.io/lidRbook/
rgl: https://cran.r-project.org/web/packages/rgl/vignettes/rgl.html
patchwork: https://cran.r-project.org/web/packages/patchwork/vignettes/patchwork.html

## Applications
* Select field sites for macrofauna ringtest: plotting
* Vegetation type classification for Global Oribatida: 
* Conservation evaluation for Soil BON Foodweb

## Concepts
.laz: compressed version of a LAS file, used to store LiDAR data.
.shp: vector file format storage developed by ESRI

## Data

las_area1 and las_area2 as example data file

```{r}
las <- readLAS("las_area1.laz")
las

plot(las)
plot(las@data$X, las@data$Y, col = las@data$Z, pch = 20)

las_check(las) # 1 points are duplicated, which is not bad

las <- readLAS("las_area1.laz", filter = "-drop_z_above 900")

#-keep_first	Keep only the first return of each pulse
#-keep_last	Keep only the last return
#-keep_single	Keep returns from pulses with only one return
#-keep_intermediate	Keep returns that are neither first nor last
#-drop_first	Exclude first returns
#-drop_last	Exclude last returns
#-drop_z_below x
#-drop_z_above 900 NB is m above sea level

plot(las, bg="white", color="Classification")

las@data$Classification %>% unique
las %>% str
las@data %>% str
```


```{r}
# exercise 2
#1. reading the point cloud
#read LAS point cloud
las<-readLAS("las_area2.laz")
las#plot the point cloud
plot(las)#read only x,y,z
# density: 4.76 pulses/m²  Pulse density (or "pulse rate") This is the number of laser pulses emitted per square meter.Pulses = quanti colpi sparo
# 8.87 points/m² number of LiDAR returns (points) per square meter. Points = quanti rimbalzi ricevo

las_check(las) #check if the point cloud is fine

#2.1 plot the point cloud
plot(las, bg="white")
plot(las, color="RGB")#if you have structrue from motion data you can plot in RGB
plot(las, color="Classification", bg="white", legend=TRUE )#plot specific information of the point cloud e.g. classification 
plot(las, color="Intensity", bg="white", legend=TRUE ) #plot intensity

plot(las@data$X, las@data$Y, col = las@data$Z, pch = 20)
```

```{r}
#exercise 1
#open the las2
#plot the point cloud
#how is the pulse density?
#check if the point cloud is fine
#re-load the point cloud and plot it using only points below  1750
#plot specific information of the point cloud of the area above 1900m: try to plot the number of returns - never done but you can try! do it with a blue background and adding a legend 
las2<-readLAS("C:/Users/MicTorresani/OneDrive - Scientific Network South Tyrol/varie/corsi/physalia/LiDAR 2025/data/starting_data/las/area2/DownloadService/las_area2.laz")
plot(las2)
las_check(las2)
las2<-readLAS("C:/Users/MicTorresani/OneDrive - Scientific Network South Tyrol/varie/corsi/physalia/LiDAR 2025/data/starting_data/las/area2/DownloadService/las_area2.laz",filter = "-drop_z_above 1750" ) 
plot(las2)
las2<-readLAS("C:/Users/MicTorresani/OneDrive - Scientific Network South Tyrol/varie/corsi/physalia/LiDAR 2025/data/starting_data/las/area2/DownloadService/las_area2.laz",filter = "-drop_z_below 1900" ) 
plot(las2, color="ReturnNumber", bg="blue", legend=TRUE )
```


```{r}
#open Qgis
las<-readLAS("las_area1.laz")
p1 <- c(679368, 5155974)
p2 <- c(679468, 5156074)
transect<-clip_transect(las, p1, p2, width = 5) #plot point cloud transect; width in meters

#transect <- las
ggplot(payload(transect), aes(X, Z, color=Z))+ #you can plot also other infos color=Intensity, color=Classification... check doing transect$
  geom_point()+
  theme_minimal()+
  scale_color_viridis_c(option="viridis")#clip the las with the shapefile of a given polygon

plot(las)
```


## Homework2
```{r}
###
#exercise 2 to do at home
#create a transect with the new las2
#show the Classification of a two given points of transect of 10m
las2<-readLAS("las_area2.laz" ) 
p1 <- c(620207, 5156764)
p2 <- c(620257, 5156764)
transect<-clip_transect(las2, p1, p2, width = 5) #plot point cloud transect

ggplot(payload(transect), aes(X,Z, color=Z  ) )+
  geom_point()+
  theme_minimal()+
  scale_color_viridis_c(option="viridis")#clip the las with the shapefile of a given polygon
```

## 2.2 plot clipped areas, circles 
```{r}
area_test<-st_read("area_shp_test.shp")

plot(area_test)
clipped_las <- clip_roi(las, area_test)
plot(clipped_las)#clip my las with given gps point and given radius
plot(clipped_las@data$X, clipped_las@data$Y, col = clipped_las@data$Z, pch = 20)

area_circle<-clip_circle(las, x=679368, y=5155974, radius = 50)
p1 <- c(679368, 5155974)
p2 <- c(679468, 5156074)
area_circle<-clip_circle(las, x=p1[1], y=p1[2], radius = 40)
#plot(area_circle)
plot(area_circle@data$X, area_circle@data$Y, col = area_circle@data$Z, pch = 20)
```

## 3.PC classification 
PC - point cloud
```{r}
#3. classification of the point cloud
#classify point cloud, we use the function classify_ground ()
#PMF
classification1 <- classify_ground(las, algorithm = pmf(ws=5, th=3) )

transect<-clip_transect(classification1, p1, p2, width = 5)
ggplot(payload(transect), aes(X,Z, color=Classification  ) )+
  geom_point()+
  ggtitle("pmf")+
  theme_minimal()+
  scale_color_viridis_c(option="viridis")


ws<-seq(3, 12, 3)
ws
th<-seq(0.1, 2, length.out= length(ws))

classification2<-classify_ground(las, algorithm = pmf(ws=ws, th=th) )
transect<-clip_transect(classification2, p1, p2, width = 5)
ggplot(payload(transect), aes(X,Z, color=Classification  ) )+
  geom_point()+
  ggtitle("pmf_tuned")+
  theme_minimal()+
  scale_color_viridis_c(option="viridis")


#CSF
library(RCSF)

classification_csf<-classify_ground(las, algorithm = csf()) #based on Zang et al 2016. # csf does not requires any tuning.
transect<-clip_transect(classification_csf, p1, p2, width = 5) #we can change the transect width if needed
ggplot(payload(transect), aes(X, Z, color=Classification  ) )+
  geom_point()+
  ggtitle("csf")+
  theme_minimal()+
  scale_color_viridis_c(option="viridis")
```

## DTM - terrain model
Important for data normalisation

```{r}
ground <- filter_ground(las)
str(ground@data)
plot(ground@data$X, ground@data$Y, col=ground@data$NumberOfReturns, pch=20)

dtm_tin <- rasterize_terrain(las, res = 1, algorithm = tin()) # res in meters | Michele the tutor knows more than five points per m2. 
# DTM generation: Triangular irregular network; recommended (for forestry)
# DTM generation: Invert distance weighting; not recommended
# DTM generation: Krigin - most advanced; for very detail terrain

dtm_tin
plot_dtm3d(dtm_tin)
plot(dtm_tin)

# export to QGIS
raster::writeRaster(dtm_tin, "Data/dtm_tin.tiff")

# dtm with idw

dtm_idw <- rasterize_terrain(las, res = 1, algorithm = knnidw())
# DTM generation: Invert distance weighting; not recommended

dtm_idw
plot_dtm3d(dtm_idw)
raster::plot(dtm_idw)
plot(dtm_idw)

# export to QGIS
raster::writeRaster(dtm_idw, "Data/dtm_idw.tiff")

dtm_products <- terrain(dtm_tin, v = c("slope", "aspect"), unit = "degree")
plot(dtm_products)
raster::writeRaster(dtm_products$slope, "Data/slope.tiff")
```

## pointcloud normalization
```{r}
nlas_wrong <- las - dtm_tin
plot(nlas_wrong, legend = TRUE)
# it is wrong because it is not accurate at all
# it shows many minus point

nlas_right <- normalize_height(las, tin())
plot(nlas_right, legend = TRUE)

# final_filter
nlas_right_filtered <- filter_poi(nlas_right, Z > 0, Z < 40)
plot(nlas_right_filtered, legend=TRUE)
```

## DSM - Digital surface model
```{r}
dsm <- rasterize_canopy(las, res=1, algorithm = dsmtin())
plot(dsm)

chm_wrong <- dsm - dtm_tin
plot(chm_wrong)

chm_right <- rasterize_canopy(nlas_right_filtered, res=1, algorithm = dsmtin())
plot(chm_right)

writeRaster(chm_right, "Data/chm_right.tiff")
```

## tree detection and segementation
```{r}
tree_top_5 <- locate_trees(nlas_right_filtered, lmf(ws = 5)) # better use the nlas (normalised) rather than the las
# tree detection with local maximum filter LMF
# windows 5 x 5 m
tree_top_5
plot(tree_top_5)
sf::st_write(tree_top_5, "Data/tree_top_5.shp", layer_options = "SHPT=POINTZ")


tree_top_10 <- locate_trees(nlas_right_filtered, lmf(ws = 10)) 
tree_top_10
plot(tree_top_10)
sf::st_write(tree_top_10, "Data/tree_top_10.shp", layer_options = "SHPT=POINTZ")

# a function to decide 
f <- function(x) {x * 0.1 + 3} #create a function that makes a mw that increase from 3 (whan height is 0) and increase of 0.1 for each meter. you can change the 0.1 and increas eto 0.2...
heights <- seq(0,40,5)
ws <- f(heights)
plot(heights, ws, type = "l", ylim = c(0,6))

tree_top_function_right<-locate_trees(nlas_right_filtered, lmf(f)) # the function adjust the ws based on tree heights
st_write(tree_top_function_right, "Data/tree_top_function.shp", delete_layer = TRUE, layer_options = "SHPT=POINTZ")
```

## Exercise
```{r}
#exercise 6
#normalize the las2 using the tin() algorithm. 
las2
nlas2 <- normalize_height(las2, tin())
plot(nlas2, legend = TRUE)

#filter the points as done before. 
plot(nlas2@data$X, nlas2@data$Y, col=nlas2@data$Classification)
plot(nlas2@data$X, nlas2@data$Z, col=nlas2@data$Classification)
nlas2_filtered <- filter_poi(nlas2, Z > 0, Z < 45)
plot(nlas2_filtered, legend=TRUE)
plot(nlas2_filtered@data$X, nlas2_filtered@data$Z, col=nlas2_filtered@data$Classification)

#exercise 7
#create a 2 new CHM with the area2 with a spatial resolution of 2m. one using the dsmtin() and the other using the p2r(subcircle = 0.2,na.fill = tin() ) #exercise 8
chm_right2 <- rasterize_canopy(nlas2_filtered, res=2, algorithm = dsmtin())
plot(chm_right2)

writeRaster(chm_right2, "Data/chm_right2.tiff")

chm_right2.2 <- rasterize_canopy(nlas2_filtered, res=2, algorithm = p2r(subcircle = .2, na.fill = tin()))
plot(chm_right2.2)

writeRaster(chm_right2.2, "Data/chm_right2.2.tiff")


#load the area2
#create a nlas_filtered (with filter 0, 45) in the area 2
#locate trees using  the f function setting height seq(0,45,5)

# a function to decide 
f <- function(x) {x * 0.1 + 3} #create a function that makes a mw that increase from 3 (whan height is 0) and increase of 0.1 for each meter. you can change the 0.1 and increas eto 0.2...
heights <- seq(0,45,5)
ws <- f(heights)
plot(heights, ws, type = "l", ylim = c(0,8))

tree_top_function_right2.2 <- locate_trees(nlas2_filtered, lmf(f)) # the function adjust the ws based on tree heights

st_write(tree_top_function_right2.2, "Data/tree_top_function_las2.2.shp", delete_layer = TRUE, layer_options = "SHPT=POINTZ")

```

## tree segmentation
```{r}
# @Dalponte2016: CHM + function to detect tree tops
# @Li2012: No CHM needed - 3D geometric clustering
# chm_right <- rasterize_canopy(nlas_right_filtered, res=1, algorithm = dsmtin())
#tree_top_function_right<-locate_trees(nlas_right_filtered, lmf(f))
algorithms1 <- dalponte2016(chm_right, tree_top_function_right)
trees_segmented_dalponte <- segment_trees(nlas_right_filtered, algorithms1)
plot(trees_segmented_dalponte, color="treeID")

plot(trees_segmented_dalponte@data$X, trees_segmented_dalponte@data$Y, col=trees_segmented_dalponte@data$treeID)

# selecting one single tree
tree1<-filter_poi(trees_segmented_dalponte, treeID==120)
plot(tree1@data$X, tree1@data$Y, col=tree1@data$Classification)

#Li2012
algorithms2 <- li2012()
trees_segmented_li2012 <- segment_trees(nlas_right_filtered, algorithms2)
plot(trees_segmented_li2012, color="treeID")

plot(trees_segmented_li2012@data$X, trees_segmented_li2012@data$Y, col=trees_segmented_li2012@data$treeID)
```

## crown metric and shape
```{r}
crown_dalponte <- crown_metrics(trees_segmented_dalponte, func = .stdtreemetrics, geom = "convex")
# convex: more linear; concave: more rough
st_write(crown_dalponte, "Data/crown_dalponte.shp")
# cnvhll is the area of the crown in meter2
```


```{r}
crown_dalponte_all_metrics <- crown_metrics(trees_segmented_dalponte, func = .stdmetrics_z, geom = "convex")
# all metrics are extracted
crown_dalponte_all_metrics
# zmax: highest point
# z: refers to the elevation
# i: intensity
# rn: return number
# q: quantile
# a: ScanAngleRank or ScanAngle
# n: count
# p: percentage
# random forests can be powerful to classify images

# get information from a single point
# tree level matrix
test_points <- st_read("punctual_plot.shp")
metric_plot <- plot_metrics(nlas_right_filtered, func = .stdmetrics_z, radius=20, test_points)

# volume
crown_dalponte_new <- crown_metrics(trees_segmented_dalponte, func = .stdmetrics_z, geom="concave")

crown_dalponte_new$volume <- 0.5*(crown_dalponte_new$zmax^2.5)/100 
# use with caution! 
# Some sources suggest values between .03 and .07 depending on species.

plot(crown_dalponte_new)
plot(crown_dalponte_new["volume"])

# volume at pixel level
crown_dalponte_new <- pixel_metrics(trees_segmented_dalponte, func = .stdmetrics_z, 5) # 5 meter resolution
crown_dalponte_new$volume <- .5*(crown_dalponte_new$zmax^2.5)/100
plot(crown_dalponte_new)
plot(crown_dalponte_new["volume"])
```


## segment trees by only CHM
## tree location using a CHM and not a point cloud
```{r}
chm_right <- rasterize_canopy(nlas_right_filtered, res=1, algorithm = dsmtin())
tree_top_chm <- locate_trees(chm_right, lmf(f))
agorithm_tree_by_chm <- dalponte2016(chm_right, tree_top_chm)
crowns <- agorithm_tree_by_chm()
plot(crowns, col = pastel.colors(2200))
# raster shape file
```



## ctg
```{r}
# starting from 01:03:00
# https://app.slack.com/client/T08V13QPH1U/C08V8HC4T60
# https://us02web.zoom.us/rec/play/KNI5VI5rRX-c0M1OvuuzphFoV7gRadI3WO4ibh3es8kCzObO4vSWugS8Ci-1f3vus-DA9yLBx9_qhM0.QSUpvIKMuCrBMMUa?eagerLoadZvaPages=sidemenu.billing.plan_management&accessLevel=meeting&canPlayFromShare=true&from=share_recording_detail&continueMode=true&componentName=rec-play&originRequestUrl=https%3A%2F%2Fus02web.zoom.us%2Frec%2Fshare%2F2aEzfAh2DFLVL43ZZnS134OrpNC1RfFU4dAGEzldrIxE6hjwUREga67uOlcKrqpQ.BupiWFprSNaNccW_

#ctg
ctg <- readLAScatalog("ctg")
crs(ctg) <- "EPSG:25832" #reference system
plot(ctg)

#create the following folders: dtm, nlas, chm, chm2,nlas_segmented, segmented

#2.dtm
#tin
opt_output_files(ctg) <- paste0("ctg/{*}_dtm") # set folder output
ctg_dtm <- rasterize_terrain(ctg, res = 5, algorithm = tin(), overwrite=TRUE) # normalize

#3.create nlas
opt_output_files(ctg) <- paste0("ctg/nlas/", "/{*}_nlas") # set folder output
ctg_nlas <- normalize_height(ctg, tin())

#4. create CHM
ctg_nlas <- readLAScatalog("ctg/nlas")
opt_output_files(ctg_nlas) <- paste0("ctg/chm/", "/{*}_chm") # settare cartella output
crs(ctg_nlas) <- "EPSG:25832"
#ctg_nlas <- catalog_filter(ctg_nlas, ~Z > 0)
ctg_chm <- rasterize_canopy(ctg_nlas, res = 1, algorithm = dsmtin())


#5.locate trees
f <- function(x) {pmax((x * 0.1 + 3), 1)} #funzione incrementale
heights <- seq(0,40,5)
ws <- f(heights)
crs(ctg_nlas) <- "EPSG:25832"
opt_output_files(ctg_nlas) <- paste0("ctg/chm/", "/{*}_ttops") # set folder output
ctg_ttops_function <- locate_trees(ctg_nlas, lmf(f))

#6.tree segmentation and tree metric in one file
opt_output_files(ctg_nlas) <- paste0("ctg/chm2", "/chm_{*}") #set folder output
chm <- rasterize_canopy(ctg_nlas, 1, dsmtin())
opt_output_files(ctg_nlas) <- ""
ttops <- locate_trees(ctg_nlas, lmf(f), uniqueness = "bitmerge")
opt_output_files(ctg_nlas) <- paste0("ctg/nlas_segmented", "/{*}_segmented") #set folder output
# check duplicated ID
ttops[duplicated(ttops$treeID), ]
ttops <- ttops[!duplicated(ttops$treeID), ]
algo <- dalponte2016(chm, ttops)
ctg_segmented <- segment_trees(ctg_nlas, algo)

#
ctg_nlas_seg <- readLAScatalog("ctg/nlas_segmented")
opt_output_files(ctg_nlas) <- paste0("ctg/segmented", "/{*}_segmented") #set folder output
crowns_dalponte <- crown_metrics(ctg_nlas_seg, func = .stdmetrics_z, geom = "convex")
st_crs(crowns_dalponte) <- 25832
#settare cartella output
st_write(
  crowns_dalponte,
  "ctg/segmented/trees_crown_and_metrics.shp",
  delete_layer = TRUE
)

#7. metrics in one file
#at raster level
#list.files("ctg")
ctg_nlas_seg <- readLAScatalog("ctg/nlas_segmented")
stmetric_pixel <- pixel_metrics(ctg_nlas_seg, func = .stdmetrics_z, 10) # calculate all the z metrics at 10 m
crs(stmetric_pixel) <- "EPSG:25832"
#settare cartella output
writeRaster(stmetric_pixel, "ctg/metrics/metriche_z_raster.tiff")
```


## gedi data
```{r}
# google earth engine code for the download of gedi global chm (potapov 30m, lang10m, meta1m):  // =======================
#// 1. Define AOI: draw or use a fixed polygon
#// =======================
#// You can also use Map.drawingTools() to draw it interactively
#var aoi = geometry;
#Map.centerObject(aoi, 13);
#// =======================
#// 2. Load Lang et al. CHM (10m)
#// =======================
var langCHM = ee.Image('users/nlang/ETH_GlobalCanopyHeight_2020_10m_v1') .clip(aoi);

Map.addLayer(langCHM, {min: 0, max: 50, palette: ['white', 'green', 'darkgreen']}, 'Lang CHM (10m)');

# 3. Load Potapov et al. CHM (GEDI V2.7) (30m)

var potapovCHM = ee.ImageCollection('users/potapovpeter/GEDI_V27')
                    .mosaic()
                    .clip(aoi);
Map.addLayer(potapovCHM, {min: 0, max: 50, palette: ['white', 'orange', 'brown']}, 'Potapov CHM (30m)');
// =======================
// 4. Load Meta CHM (1m) if available
// =======================
var metaCHM = ee.ImageCollection("projects/meta-forest-monitoring-okw37/assets/CanopyHeight")
                .filterBounds(aoi)
                .mosaic()
                .clip(aoi);
Map.addLayer(metaCHM, {min: 0, max: 50, palette: ['white', 'blue', 'purple']}, 'Meta CHM (1m)');
// =======================
// 5. Export to Google Drive
// =======================
// Lang export (10 m)
Export.image.toDrive({
  image: langCHM,
  description: 'Lang_CHM_10m',
  folder: 'GEE_exports',
  region: aoi,
  scale: 10,
  maxPixels: 1e13
});
// Potapov export (30 m)
Export.image.toDrive({
  image: potapovCHM,
  description: 'Potapov_CHM_30m',
  folder: 'GEE_exports',
  region: aoi,
  scale: 30,
  maxPixels: 1e13
});
// Meta export (1 m)
Export.image.toDrive({
  image: metaCHM,
  description: 'Meta_CHM_1m',
  folder: 'GEE_exports',
  region: aoi,
  scale: 1,
  maxPixels: 1e13
});
```


## Older codes
## The LidR
```{r}
dev.new()
rgl::rgl.open()
dev.off()
LASfile <- system.file("extdata", "Topography.laz", package="lidR")
las <- readLAS(LASfile, select = "xyzc")
# plot(las, size = 3, bg = "white")
plot(las@data$X, las@data$Y, col = las@data$Z, pch = 20)

dtm_tin <- rasterize_terrain(las, res = 1, algorithm = tin())
plot_dtm3d(dtm_tin, bg = "white") 

gnd <- filter_ground(las)
plot(gnd, size = 3, bg = "white", color="Classification")

dtm <- rasterize_terrain(las, 1, knnidw())
plot(dtm, col = gray(1:50/50))

# normalization point cloud by - surface
nlas <- las - dtm
plot(nlas, size = 4, bg = "white")

hist(filter_ground(nlas)$Z, breaks = seq(-0.6, 0.6, 0.01), main = "", xlab = "Elevation")
```

```{r}
chm <- rasterize_canopy(las, res = 3, algorithm = p2r())
col <- height.colors(25)
plot(chm, col = col)

LASfile <- system.file("extdata", "Topography.laz", package = "lidR")
las2 <- readLAS(LASfile)
las2 <- normalize_height(las2, algorithm = tin())
plot(las2, size = 3, bg = "white")
chm <- rasterize_canopy(las2, res = 0.5, algorithm = dsmtin())
plot(chm, col = col)

chm <- rasterize_canopy(las2, res = .5, algorithm = dsmtin(max_edge = 8))
plot(chm, col = col)
```

```{r}
# Individual tree detection and segementation
LASfile <- system.file("extdata", "MixedConifer.laz", package="lidR")
las <- readLAS(LASfile, select = "xyzr", filter = "-drop_z_below 0")
chm <- rasterize_canopy(las, 0.5, pitfree(subcircle = 0.2))
plot(las, bg = "white", size = 4)
plot(las@data$X, las@data$Y, col = las@data$Z, pch = 20)

# local maximum filter with fixed window size
ttops <- locate_trees(las, lmf(ws = 5))

plot(chm, col = height.colors(50))
plot(sf::st_geometry(ttops), add = TRUE, pch = 3)

x <- plot(las, bg = "white", size = 4)
add_treetops3d(x, ttops)

```

```{r}
# derived matrics at the tree level
LASfile <- system.file("extdata", "MixedConifer.laz", package="lidR") 
las <- readLAS(LASfile, filter = "-drop_z_below 0") # read the file
metrics <- crown_metrics(las, ~list(z_max = max(Z), z_mean = mean(Z))) # calculate tree metrics
head(metrics)

plot(metrics["z_max"], pal = hcl.colors, pch = 19) # plot using z_max

custom_crown_metrics <- function(z, i) { # user-defined function
  metrics <- list(
     z_max = max(z),   # max height
     z_sd = sd(z),     # vertical variability of points
     i_mean = mean(i), # mean intensity
     i_max  = max(i)   # max intensity
   )
   return(metrics) # output
}

ccm = ~custom_crown_metrics(z = Z, i = Intensity)

metrics <- crown_metrics(las, func = ccm, geom = "convex")
plot(metrics["z_max"], pal = hcl.colors)

metrics <- crown_metrics(las, func = ccm, geom = "concave")
plot(metrics["z_max"], pal = hcl.colors)
```


## 

